Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-04-23 overview.tex
Finished editing section 2.

\subsection{Physical Equations}

@@

\subsection{Overview of Numerical Methods}

In this section, we briefly describe the numerical methods which are used to solve the equations outlined above.  We sweep through in the same order as will be used in section~\ref{sec.methods} so that there is a one-to-one correspondence between each of the following overviews and the complete description provided in that section.  The goal here is to introduce the reader to the basic idea of the methods without drowning in detail.

@@

The primary purpose of the \enzo\ code is its Adaptive Mesh Refinement
capability, which allows it to reach extremely large dynamical ranges
with limited computational resources, opening doors previously closed
by finite memory and computational time. Unlike moving mesh methods
\citep{1995ApJS..100..269P,1995ApJS...97..231G} or  
methods that subdivide 
individual cells \citep{Adjerid}, Berger \& Collela's AMR (also referred 
to as \emph{structured} AMR) utilizes an adaptive hierarchy of grid 
patches at varying levels of resolution.  Each rectangular grid patch 
(referred to as a ``grid'') covers some region of space in its 
\emph{parent grid} which requires higher resolution, and can itself 
become the parent grid to an even more highly resolved \emph{child grid}.

The grid hierarchy begins with the root grid which covers the entire
domain of interest with a coarse, uniform, Cartesian grid. Then, as
the solution evolves and interesting regions form, finer meshes are
placed below these regions (we use the notation `below' to refer to
finer grids and `above' for coarser grids).  We restrict the ratio
between cell sizes to be an integer, typically 2-4, and refer to a
level as all the grids with the same cell size.  In order to keep
things as simple as possible, the edges of subgrids must coincide with
the cell edge of its immediate parent (coarser) grid. Additionally,
the hierarchy can be initialized with one or more static grids if a
higher initial resolution is required.

Given the hierarchy at some time $t$, we advance the solution in the
manner of a W-cycle in a multigrid solver.  First, we determine the
maximum time step allowed for the coarsest grid based on a variety of
accuracy and stability criteria and advance the grid by that time
interval, $\Delta t_0$.  We then move down to the next level and
advance all the grids on that level by a timestep $\Delta t_1$
($\Delta t_1 \leq \Delta t_0$) which is the minimum of all the allowed
timesteps for those grids.  If there are more levels, we repeat this
procedure until the bottom level of the hierarchy has been reached.
Once there, we continue advancing the grids on the lowest level until
they have `caught up' to the next highest level above (i.e. $\sum
\Delta t_l = \Delta t_{l-1}$).  This procedure repeats itself until
all grids have been advanced by a total time of $\Delta t_0$.

Since interesting regions on the grid may move, the hierarchy must
adapt itself.  We do this whenever a level has caught up to the
coarser level above it and consists of entirely rebuilding the grids
on that level and below.  This is done by applying the grid refinement
criterion to the grids on that level, flagging zones which require
extra grids.  This criterion depends on the physical problem being
simulated.  We have implemented a number of options, including shock
and steep gradient detectors, but in the problem described here,
employ one based on the mass within a cell, imitating the Lagrangian
nature of the SPH algorithm.  Once a grid has a set of flagged cells,
we run a machine-vision based algorithm \citep{Berger91} to find edges
and determine a good placement of subgrids.  These subgrids must not
overlap one another, must cover all flagged cells and their
neighbouring cells, and be above a preset efficiency threshold, where
the efficiency of a cell is defined as the ratio of flagged cells to
total cells.  Once these new subgrids have been identified, the
solution from the next coarser grid is interpolated (see below) in
order to initialize the values on the new grids.  Finally, any overlap
between these new subgrids and the old ones is identified and the
solution within the regions of overlap is copied to the new subgrids.
The entire procedure just outlined is then repeated on the new grids
and in this way the entire hierarchy (from the original level examined
and below) is rebuilt.

@@

Four different (magneto)-hydrodynamic methods are implemented in \enzo: (i) the
HD-only piecewise parabolic method (PPM) developed by~\citet{1984JCoPh..54..174C}
and extended to cosmology by~\citet{1995CoPhC..89..149B};
(ii) the MUSCL-like Gudonov scheme \citep{MUSCL} with or without magnetic fields
and Dedner-based divergence cleansing and described in more detail in 
\citet{WangAbelZhang08} and \citet{WangAbel09};
(iii) a constrained transport staggered MHD scheme \citet{Collins10}, and
(iv) the second-order finite-difference hydrodynamics method described 
in \zeus~\citep{Stone92a,Stone92b}.

We begin with the direct-Eulerian PPM implementation.  
This is an explicit, higher order-accurate version of
Godunov's method for ideal gas dynamics with third order-accurate piecewise parabolic
monotonic interpolation and a nonlinear Riemann solver for shock
capturing.  It advances the hydrodynamic equations in the following
steps: 
%\begin{enumerate}
% \item 
 (i) Construct monotonic parabolic interpolation of cell average data, for each fluid quantity.
% \item 
(ii) Compute interface states by averaging the parabola for over the domain of dependence for each interface
% \item 
(iii) Use interface data to solve the Riemann problem.
 %\item 
 (iv) Difference the interface fluxes to update the cell average quantities.
%\end{enumerate}

It does an excellent job of capturing strong shocks in a few cells; Multidimensional schemes are built up by directional
splitting and produce a method that is formally second order-accurate
in space and time which explicitly conserves mass, linear momentum, and energy.  A variety of
reconstruction methods and Riemann solvers have been implemented.

As described in \citet{Bryan95}, we modify the method for use in hypersonic flows when the thermal energy $e$ is extremely small compared to the total energy $E$.  This is a problem because in total energy method, the temperature is computed by subtracting one large number from another (i.e. the kinetic energy from the total energy), a situation which is can generate large numerical inaccuracies.  We address this situation by also solving the thermal energy equation and using $e$ from this equation when we expect the error to be large.

@@

The last hydrodynamics method we briefly describe is the finite-different algorithm used in the ZEUS, as given in more detai in \citet{Stone92a}.  For this reason, we call this the ZEUS method, although the code is entirely independent of the ZEUS code, and only the hydrodynamical algorithm of \zeus\ is implemented in \enzo -- the MHD and RHD schemes are not.

The ZEUS method uses a staggered mesh such that the velocities are face-centered, while the density and thermal energy quantities are cell-centered.  It splits the solution up into two steps, the so-called source step, in which the momentum and energy values are updated to reflect the pressure and gravity forces, as well as the effect of an artificial viscosity required for stability.  The transport step accounts for the advection of conserved quantities (mass, momentum and energy) across the grid.

@@

The current implementation of self-gravity in \enzo uses a Fast Fourier Technique \citep{Hockney88} to solve Poisson's equation on the top (level 0) grid.  The advantage of this is that it naturally allows both periodic and isolated boundary conditions for the gravity, choices which are very common in astrophysics and cosmology.  On subgrids, we interpolate the boundary conditions from that grid's parent (either the level 0 grid or some other subgrid).  The Poisson equation is then solved using a multigrid technique.  In \enzo\, self-gravity is optional.  There are also a number of options for static gravitational fields.

@@

Collisionless matter (dark matter, stars, etc.) is modeled with particles which interact with the baryons only via gravity.  These particles are advanced in a single time step using a drift-kick-drift algorithm to provide second-order accuracy even in the presence of changing timesteps.  Since the particles follow the collapse of structure, they are not adaptively refined.  Nor are there duplicate sets of particles for each level; instead, each particle is associated with the highest refined
level available at its position in space, and moved as the hierarchy is rebuilt.  Thus, a particle has the same timestep and feels the same gravitational force as the grid at that level.

@@

In addition to this, \enzo\ includes the capability of following up to 12 chemical species using a non equilibrium-solver.  The species can be turned on in sets, with the simplest model including just H, H$^+$, He, He$^+$, He$^++$, and $e^-$, and more complete models adding first species important for gas-phase molecular hydrogen formation: $H^-$, H$_2$ and H$_2^+$, and then $HD$ formation: $HD$, $D$, $D+$.  The cooling and heating due to these species is also included (see the next section).  The solution of the rate equations is carried out using a time-backwards differencing which insures stability; to maintain accuracy we sub-cycle the rate equations with a shorter timestep such that the electron and neutral fractions do not change by more than 10\%.

@@

Enzo may operate in a number of modes with regard to radiative cooling and heating.  In the simplest mode, with the multi-species flag turned off (i.e. set to zero), the cooling rate is computed from a simple temperature-dependent cooling rate, taken from \citet{SW87}.  If chemistry is turned on, then the code can include cooling from all species of H and He (including H$_2$) -- this primordial cooling is also treated using a backwards-time differencing scheme with sub-cycling to maintain accuracy.  Finally, we have also recently added metal cooling based on a set of multi-dimensional lookup tables computed with the \texttt{Cloudy} code \citep{1998PASP..110..761F} as described in \citet{2008MNRAS.385.1443S} and \citet{2011ApJ...731....6S}.  Finally, we note that the cooling and heating is (almost always) treated in the optically-thin limit, although the code can also follow radiative transfer in a limited set of energy bins (see below).

@@

The chemical networks and heating rates can be affected by external radiation fields, and the code includes a number of pre-set radiation backgrounds that are uniform in space but can very in time.  These are generally based on the redshift-dependent rates given in \citet{1996ApJ...461...20H} and \citet{2012ApJ...746..125H}, but can also include a uniform H$_2$ photo-dissociating background, either constant in time or varying as in \citet{WiseAbel05}.

@@

The other option for radiative transfer is a moment-based method which adds an addition field describing the energy-density of radiation.  This field is evolved using the flux-limited diffusion method which transitions smoothly between streaming (optically thin) and opaque limits and is coupled to the hydrogen ionization network.  The resulting set of linear equations is solved using the parallel HYPRE framework.  Full details for the enzo implementation can be found in \citet{ReynoldsHayesPaschosNorman2009}.

@@

Heat conduction, both isotropic and anisotropic, can be included using an sub-cycled, operator-split method.  The heat fluxes are computed with simple second-order accurate finite differences and stability is ensured by restricting the time step, and using flux-limiters.  This is described in more detail below.

@@

A simple heuristic method is used to model the formation of stars and their feedback of metals and energy into the gas.  This method is based on \citep{CO1992} and involves identify plausible cites of star formation based on a set of criteria (dense gas with a short cooling time, which is both collapsing and unstable), and computing a star formation rate based on the Schmidt-Kennicutt relation \citep{K89}.  The relating gas is converted into a star particle over a few dynamical times, and over this timespan metals and thermal energy are injected into the single, local cell occupied by the star particle.

>>>>

\subsection{Physical Equations}
\label{sec.equations}

@@

\subsection{Overview of Numerical Methods}
\label{sec.method_overview}

In this section, we briefly describe the numerical methods which are
used to solve the equations outlined in Section~\ref{sec.equations}.  We
proceed through the numerical methods in the same order as will be used in
section~\ref{sec.methods} so that there is a one-to-one correspondence
between each of the following overviews and the complete description
provided in that section.  The goal here is to introduce the reader to
the basic principles of the methods without drowning in detail.

@@

The primary purpose of the \enzo\ code is its Adaptive Mesh Refinement (AMR)
capability, which allows it to reach extremely large dynamical ranges
with limited computational resources, opening doors previously closed
by finite memory and computational time. Unlike moving mesh methods
\citep{1995ApJS..100..269P,1995ApJS...97..231G} or  
methods that subdivide 
individual cells \citep{Adjerid}, Berger \& Collela's AMR (also referred 
to as \emph{structured} AMR) utilizes an adaptive hierarchy of grid 
patches at varying levels of resolution.  Each rectangular grid patch 
(referred to as a ``grid'') covers some region of space in its 
\emph{parent grid} which requires higher resolution, and can itself 
become the parent grid to an even more highly resolved \emph{child grid}.

The grid hierarchy begins with the root grid, which covers the entire
domain of interest with a coarse, uniform, Cartesian grid. Then, as
the solution evolves and interesting regions form, finer meshes are
placed below these regions (we use the notation `below' to refer to
finer grids and `above' for coarser grids).  We restrict the ratio
between cell sizes to be an integer, typically 2 or 4, and refer to a
level as all grids with the same cell size.  In order to keep
things as simple as possible, the edges of subgrids must coincide with
the cell edge of its immediate parent (coarser) grid. Additionally,
the hierarchy can be initialized with one or more static grids if a
higher initial resolution is required.

Given the hierarchy at some time $t$, we advance the solution in the
manner of a W-cycle in a multigrid solver.  First, we determine the
maximum time step allowed for the coarsest grid based on a variety of
accuracy and stability criteria and advance the grid by that time
interval, $\Delta t_0$.  We then move down to the next level and
advance all the grids on that level by a timestep $\Delta t_1$
($\Delta t_1 \leq \Delta t_0$), which is the minimum of all the allowed
timesteps for those grids.  If there are more levels, we repeat this
procedure until the bottom level of the hierarchy has been reached.
Once there, we continue advancing the grids on the lowest level until
they have `caught up' to the next highest level above (i.e. $\sum
\Delta t_l = \Delta t_{l-1}$).  This procedure repeats itself until
all grids have been advanced by a total time of $\Delta t_0$.

Since interesting regions on the grid may move, the hierarchy must
adapt itself.  We do this whenever a level has caught up to the
coarser level above it and consists of entirely rebuilding the grids
on that level and below.  This is done by applying the grid refinement
criteria to the grids on that level, flagging zones that require
extra grids.  These criteria depend on the physical problem being
simulated.  We have implemented a number of options, including shock
and steep gradient detectors, but in the problem described here,
employ one based on the mass within a cell, imitating the Lagrangian
nature of the SPH algorithm.  Once a grid has a set of flagged cells,
we run a machine-vision based algorithm \citep{Berger91} to find edges
and determine a good placement of subgrids.  These subgrids must not
overlap one another, must cover all flagged cells and their
neighbouring cells, and be above a preset efficiency threshold, where
the efficiency of a cell is defined as the ratio of flagged cells to
total cells.  Once these new subgrids have been identified, the
solution from the next coarser grid is interpolated (see below) in
order to initialize the values on the new grids.  Finally, any overlap
between these new subgrids and the old ones is identified and the
solution within the regions of overlap is copied to the new subgrids.
The entire procedure just outlined is then repeated on the new grids
and in this way the entire hierarchy (from the original level examined
and below) is rebuilt.

@@

Four different (magneto)-hydrodynamic methods are implemented in
\enzo: (i) the hydrodynamic-only piecewise parabolic method (PPM)
developed by~\citet{1984JCoPh..54..174C} and extended to cosmology
by~\citet{1995CoPhC..89..149B}; (ii) the MUSCL-like Gudonov scheme
\citep{MUSCL} with or without magnetic fields and Dedner-based
divergence cleaning, described in more detail in
\citet{WangAbelZhang08} and \citet{WangAbel09}; (iii) a constrained
transport staggered MHD scheme \citep{Collins10}, and (iv) the
second-order finite-difference hydrodynamics method described in
\zeus~\citep{Stone92a,Stone92b}.

We begin with the direct-Eulerian PPM implementation.  This is an
explicit, higher-order-accurate version of Godunov's method for ideal
gas dynamics with third order-accurate piecewise parabolic monotonic
interpolation and a nonlinear Riemann solver for shock capturing.  It
advances the hydrodynamic equations in the following steps:
%\begin{enumerate}
% \item 
 (i) Construct monotonic parabolic interpolation of cell average data, for each fluid quantity.
% \item 
(ii) Compute interface states by averaging the parabola over the domain of dependence for each interface
% \item 
(iii) Use interface data to solve the Riemann problem.
 %\item 
 (iv) Difference the interface fluxes to update the cell average quantities.
%\end{enumerate}

This PPM implementation does an excellent job of capturing strong
shocks in a few cells.  Multidimensional schemes are built up by
directional splitting and produce a method that is formally second
order-accurate in space and time and which explicitly conserves mass,
linear momentum, and energy.  A variety of reconstruction methods and
Riemann solvers have been implemented.

As described in \citet{Bryan95}, we modify the method for use in
hypersonic flows when the thermal energy $e$ is extremely small
compared to the total energy $E$.  This is a problem because in the
total energy method, the temperature is computed by subtracting one
large number from another (i.e. the kinetic energy from the total
energy), a situation which can generate large numerical inaccuracies.
We address this situation by also solving the thermal energy equation
and using $e$ from this equation when we expect the error to be large.

@@

The last hydrodynamics method we briefly describe is the
finite-difference algorithm used in the \zeus\ code, as given in more
detail in \citet{Stone92a}.  For this reason, we call this the \zeus\
method, although the code is entirely independent of the \zeus\ code,
and only the hydrodynamical algorithm of \zeus\ is implemented in
\enzo -- the MHD and RHD schemes are not.

The ZEUS method uses a staggered mesh such that the velocities are
face-centered, while the density and internal energy quantities are
cell-centered.  It splits the solution up into two steps, the
so-called source step, in which the momentum and energy values are
updated to reflect the pressure and gravity forces, as well as the
effect of an artificial viscosity required for stability.  The second
step (known as the transport step) accounts for the advection of
conserved quantities (mass, momentum and energy) across the grid.

@@

The current implementation of self-gravity in \enzo\ uses a Fast
Fourier Technique \citep{Hockney88} to solve Poisson's equation on the
top (level 0) grid.  The advantage of using this method is that it
naturally allows both periodic and isolated boundary conditions for
the gravity, choices which are very common in astrophysics and
cosmology.  On subgrids, we interpolate the boundary conditions from
that grid's parent (either the level 0 grid or some other subgrid).
The Poisson equation is then solved using a multigrid technique.  In
\enzo, self-gravity is optional.  Aside from calculating the
gravitational potential from the baryon fields and particles existing
in the simulation, there are also a number of options for static gravitational fields.

@@

Collisionless matter (dark matter, stars, etc.) is modeled with
particles that interact with the baryons only via gravity.  These
particles are advanced in a single time step using a drift-kick-drift
algorithm to provide second-order accuracy even in the presence of
changing timesteps.  Since the particles follow the collapse of
structure, they are not adaptively refined.  Nor are there duplicate
sets of particles for each level; instead, each particle is associated
with the highest refined level available at its position in space, and is
moved as the hierarchy is rebuilt.  Thus, a particle has the same
timestep and feels the same gravitational force as a grid at that refinement
level.

@@

\enzo\ includes the capability of following up to
12 chemical species using a non-equilibrium solver.  The species can
be turned on in sets, with the simplest model including just H, H$^+$,
He, He$^+$, He$^++$, and $e^-$, and more complete models adding first
species important for gas-phase molecular hydrogen formation: $H^-$,
H$_2$ and H$_2^+$, and then $HD$ formation: $HD$, $D$, $D+$.  The
cooling and heating due to these species is also included (see the
next section).  The solution of the rate equations is carried out
using a time-backwards differencing which insures stability; to
maintain accuracy we sub-cycle the rate equations with a shorter
timestep such that the electron and neutral fractions do not change by
more than 10\%.

@@

Enzo may operate in a number of modes with regard to radiative cooling
and heating.  In the simplest mode, with the multi-species flag turned
off (i.e. set to zero), the cooling rate is computed from a simple
temperature-dependent cooling rate, taken from \citet{SW87}.  If
chemistry is turned on, then the code can include cooling from all
species of Hydrogen and Helium (including H$_2$) -- this primordial cooling is
also treated using a backwards-time differencing scheme with
sub-cycling to maintain accuracy.  Finally, we have also recently
added metal cooling based on a set of multi-dimensional lookup tables
computed with the \texttt{Cloudy} code \citep{1998PASP..110..761F} as
described in \citet{2008MNRAS.385.1443S} and
\citet{2011ApJ...731....6S}.  Finally, we note that the cooling and
heating is (almost always) treated in the optically-thin limit,
although the code can also follow radiative transfer in a limited set
of energy bins (see below).

@@

The chemical networks and heating rates described in the previous
section can be affected by external radiation fields, and the code
includes a number of pre-set radiation backgrounds that are uniform in
space but can very in time.  These are generally based on the
redshift-dependent rates given in \citet{1996ApJ...461...20H} and
\citet{2012ApJ...746..125H}, but can also include a uniform H$_2$
photo-dissociating background that is either constant in time or
varying as in \citet{WiseAbel05}.

@@

A second option for radiative transfer is a moment-based method that
adds an additional field tracking the radiant energy density
This field is evolved using the flux-limited diffusion method, which
transitions smoothly between streaming (optically thin) and opaque
limits and is coupled to the hydrogen ionization network.  The
resulting set of linear equations is solved using the parallel HYPRE
framework.  Full details for the enzo implementation can be found in
\citet{ReynoldsHayesPaschosNorman2009}.

@@

Heat conduction, both isotropic and anisotropic, can be included using
an sub-cycled, operator-split method.  The heat fluxes are computed
with simple second-order accurate finite differences and stability is
ensured by restricting the time step, and using flux-limiters where
appropriate.

@@

A family of simple heuristic methods are used to model the formation
of stars and their feedback of metals and energy into the gas.  These
methods are based on the work of \citet{CO1992}, and involve the
identification of plausible sites of star formation based on a set of
criteria (for example, dense gas with a short cooling time, which is
both collapsing and unstable).  The local star formation rate is
computed using a range of methods, such as a density-dependent method
based on the Schmidt-Kennicutt relation \citep{K89}.  The affected gas
is converted into a star particle over a few dynamical times, and
metals and thermal energy are injected into the region surrounding the
star particle.
